Verification of a Deterministic Random Bits Generator

### 2015-12-23 paper/paper.tex
correct spelling of couriers package
changed color of citations and ref
paper: appendix B

\usepackage{courier}
\usepackage[normalem]{ulem}
\usepackage{listings}
\usepackage{lstlangcoq}
\usepackage{appendix}

\lstset{language=Coq,basicstyle=\ttfamily,mathescape=true,columns=fullflexible,breaklines=true,breakatwhitespace=true}

@@

\begin{lstlisting}[language=C]
    int mbedtls_md_hmac_reset( mbedtls_md_context_t *ctx ) {
        HMAC_CTX * hmac_ctx = ctx->hmac_ctx;
        HMAC_Init(hmac_ctx, NULL, SHA256_DIGEST_LENGTH);
        return 0;
    }
\end{lstlisting}
\section{DRBG API Specifications} \label{appendix_api}

>>>>

\usepackage{hyperref}
\usepackage{couriers}
\usepackage[normalem]{ulem}
\usepackage{listings}
\usepackage{lstlangcoq}
\usepackage{appendix}

\lstset{language=Coq,basicstyle=\ttfamily,mathescape=true,columns=fullflexible,breaklines=true,breakatwhitespace=true}
\hypersetup{citecolor=black,linkcolor=black}

@@

\begin{lstlisting}[language=C]
    int mbedtls_md_hmac_reset( mbedtls_md_context_t *ctx ) {
        HMAC_CTX * hmac_ctx = ctx->hmac_ctx;
        HMAC_Init(hmac_ctx, NULL, SHA256_DIGEST_LENGTH);
        return 0;
    }
\end{lstlisting}
\section{DRBG API Specifications} \label{appendix_api}
In this section, I present a detailed description of the API specifications for the following functions: \lstinline{update}, \lstinline{reseed}, and \lstinline{generate}.

\subsection{The Update Function}

The API specification is as follows:

\begin{lstlisting}
    Definition hmac_drbg_update_spec :=
      DECLARE _mbedtls_hmac_drbg_update
       WITH contents: list Z,
            additional: val, add_len: Z,
            ctx: val, initial_state: hmac256drbgstate,
            initial_state_abs: hmac256drbgabs,
            kv: val, info_contents: md_info_state
        PRE [ _ctx OF (tptr t_struct_hmac256drbg_context_st), _additional OF (tptr tuchar), _add_len OF tuint ]
           PROP (
             0 <= add_len <= Int.max_unsigned;
             Zlength (hmac256drbgabs_value initial_state_abs) = Z.of_nat SHA256.DigestLength;
             add_len = Zlength contents;
             Forall isbyteZ (hmac256drbgabs_value initial_state_abs);
             Forall isbyteZ contents
           )
           LOCAL (temp _ctx ctx; temp _additional additional; temp _add_len (Vint (Int.repr add_len)); gvar sha._K256 kv)
           SEP (
             (data_at Tsh (tarray tuchar add_len) (map Vint (map Int.repr contents)) additional);
             (data_at Tsh t_struct_hmac256drbg_context_st initial_state ctx);
             (hmac256drbg_relate initial_state_abs initial_state);
             (data_at Tsh t_struct_mbedtls_md_info info_contents (hmac256drbgstate_md_info_pointer initial_state));
             (K_vector kv)
               )
        POST [ tvoid ]
           PROP ()
           LOCAL ()
           SEP (
             (hmac256drbgabs_common_mpreds (hmac256drbgabs_hmac_drbg_update initial_state_abs contents) initial_state ctx info_contents);
             (data_at Tsh (tarray tuchar add_len) (map Vint (map Int.repr contents)) additional);
             (K_vector kv)
           ).
\end{lstlisting}

The first \lstinline{WITH} variable is \lstinline{contents}, which is mathematical representation of the sequence of bytes to be consumed by the update function (\lstinline{provided_data} in the functional specification). The next two \lstinline{WITH} values are the pointer value, \lstinline{additional}, at which \lstinline{contents} is located in memory, and the length of \lstinline{contents}, \lstinline{add_len}. In the PROP part of the specification, \lstinline{add_len} is restricted to be a valid unsigned integer, and the length of \lstinline{contents} is enforced to be \lstinline{add_len}. The mathematical integers in \lstinline{contents} are also enforced to be valid bytes. In the \lstinline{SEP} part, \lstinline{additional} is said to point to \lstinline{contents}. In the postcondition, the same SEP clause appears, showing that the update function does not change the memory pointed to by \lstinline{additional}.

The next \lstinline{WITH} value is the pointer value, \lstinline{ctx}, pointing to the DRBG struct. The next two \lstinline{WITH} values are the abstract and concrete DRBG states, respectively. Skipping the next \lstinline{WITH} value, \lstinline{kv}, for the moment, the last \lstinline{WITH} value is the content of the struct denoting what kind of HMAC is being used, \lstinline{info_contents}. In the \lstinline{PROP} clause, the sequence \lstinline{V} part of the abstract state is enforced to be the size of the output of the HMAC function and enforced to contain valid bytes. This is an invariant that the update function and all public functions enforce, and so will be seen throughout all API specifications. In the \lstinline{SEP} clause, \lstinline{ctx} is said to point to an area in memory that contains the concrete state, and the concrete state is related to the abstract state by the equivalence definition. The info pointer from the concrete state also points to a valid address containing \lstinline{info_contents}. In the post condition, there is a new abstract state calculated by \lstinline{hmac256drbgabs_hmac_drbg_update}, which is just a wrapper around the functional specification with the HMAC256 function used. In the \lstinline{SEP} part, there exists a corresponding concrete state pointed to by \lstinline{ctx} that is equivalent to the new abstract state. This new concrete state’s info pointer also points to the same valid memory location containing \lstinline{info_contents}. This is what the definition \lstinline{hmac256drbgabs_common_mpreds} encapsulates.

Lastly, the \lstinline{WITH} value \lstinline{kv} represents the pointer value to the SHA K vector, which is a global constant array in the SHA-256 implementation. For more information, see the \cite{sha}.

\subsection{The Reseed Function}

The API specification for the \lstinline{reseed} function is as follows:

\begin{lstlisting}
    Definition hmac_drbg_reseed_spec :=
      DECLARE _mbedtls_hmac_drbg_reseed
       WITH contents: list Z,
            additional: val, add_len: Z,
            ctx: val, initial_state: hmac256drbgstate,
            initial_state_abs: hmac256drbgabs,
            kv: val, info_contents: md_info_state,
            s: ENTROPY.stream
        PRE [ _ctx OF (tptr t_struct_hmac256drbg_context_st), _additional OF (tptr tuchar), _len OF tuint ]
           PROP (
             0 <= add_len <= Int.max_unsigned;
             Zlength (hmac256drbgabs_value initial_state_abs) = Z.of_nat SHA256.DigestLength;
             add_len = Zlength contents;
             hmac256drbgabs_entropy_len initial_state_abs = 32;
             Forall isbyteZ (hmac256drbgabs_value initial_state_abs);
             Forall isbyteZ contents
           )
           LOCAL (temp _ctx ctx; temp _additional additional; temp _len (Vint (Int.repr add_len)); gvar sha._K256 kv)
           SEP (
             (data_at Tsh (tarray tuchar add_len) (map Vint (map Int.repr contents)) additional);
             (data_at Tsh t_struct_hmac256drbg_context_st initial_state ctx);
             (hmac256drbg_relate initial_state_abs initial_state);
             (data_at Tsh t_struct_mbedtls_md_info info_contents (hmac256drbgstate_md_info_pointer initial_state));
             (Stream s);
             (K_vector kv)
               )
        POST [ tint ]
           EX ret_value:_,
           PROP (
               return_value_relate_result (mbedtls_HMAC256_DRBG_reseed_function s initial_state_abs contents) ret_value
             )
           LOCAL (temp ret_temp ret_value)**
**           SEP (
             (hmac256drbgabs_common_mpreds (hmac256drbgabs_reseed initial_state_abs s contents) initial_state ctx info_contents);
             (data_at Tsh (tarray tuchar add_len) (map Vint (map Int.repr contents)) additional);
             (Stream (get_stream_result (mbedtls_HMAC256_DRBG_reseed_function s initial_state_abs contents)));
             (K_vector kv)
           ).
\end{lstlisting}

Notice a new notation here: \lstinline{EX ret_value:_}. This can be read literally as “there exists a variable \lstinline{ret_value} that satisfies the postcondition”. This existential variable captures the return value of the function, and its value is enforced in the \lstinline{PROP} part of the postcondition. The return value is enforced to be 0 if the reseed succeeds, a specific value if the reseed fails because of a failure to get entropy, and any non-zero value if the reseed fails for any other reason.

Note that \lstinline{mbedtls_HMAC256_DRBG_reseed_function} is just a wrapper around the functional specification of the reseed function, with the reseed algorithm and other dependencies filled in. \lstinline{hmac256drbgabs_reseed} returns the resulting abstract state from the reseed function if the reseed succeeds, or the initial abstract state otherwise.

Much like the \lstinline{update} function, the first three \lstinline{WITH} values are used for the additional data parameter. \lstinline{additional} is a pointer value that points to \lstinline{contents}, of length \lstinline{add_len}, and the function leaves this relation untouched.

Again, much like the \lstinline{update} function, the next three \lstinline{WITH} values are the pointer value to the DRBG state, the abstract DRBG state, and the concrete DRBG state. The only new property is in the \lstinline{PROP} part of the precondition, requiring the \lstinline{entropy_len} component to be 32. Note that in \stdtitle{NIST SP 800-90A} and consequently in the functional specification, this is a global, fixed constant, but \stdtitle{mbedTLS} varies this value based on the type of HMAC used and fixes the value upon instantiation. With HMAC256, this is fixed to be 32.

Unlike the \lstinline{update} function, the \lstinline{reseed} function needs to reason about entropy; that is the whole point of reseeding! The \lstinline{WITH} value \lstinline{s} represents the entropy stream to be used, and a new one is passed back out in the \lstinline{SEP} part of the postcondition.

\subsection{The Generate Function}

The API specification for the \lstinline{generate} function is as follows:

\begin{lstlisting}
    Definition hmac_drbg_generate_spec :=
      DECLARE _mbedtls_hmac_drbg_random_with_add
       WITH contents: list Z,
            additional: val, add_len: Z,
            output: val, out_len: Z,
            ctx: val, initial_state: hmac256drbgstate,
            initial_state_abs: hmac256drbgabs,
            kv: val, info_contents: md_info_state,
            s: ENTROPY.stream
        PRE [ _p_rng OF (tptr tvoid), _output OF (tptr tuchar), _out_len OF tuint, _additional OF (tptr tuchar), _add_len OF tuint ]
           PROP (
             0 <= add_len <= Int.max_unsigned;
             0 <= out_len <= Int.max_unsigned;
             Zlength (hmac256drbgabs_value initial_state_abs) = Z.of_nat SHA256.DigestLength;
             add_len = Zlength contents;
             hmac256drbgabs_entropy_len initial_state_abs = 32;
             hmac256drbgabs_reseed_interval initial_state_abs = 10000;
             0 <= hmac256drbgabs_reseed_counter initial_state_abs <= Int.max_signed;
             Forall isbyteZ (hmac256drbgabs_value initial_state_abs);
             Forall isbyteZ contents
           )
           LOCAL (temp _p_rng ctx; temp _output output; temp _out_len (Vint (Int.repr out_len)); temp _additional additional; temp _add_len (Vint (Int.repr add_len)); gvar sha._K256 kv)
           SEP (
             (data_at_ Tsh (tarray tuchar out_len) output);
             (data_at Tsh (tarray tuchar add_len) (map Vint (map Int.repr contents)) additional);
             (data_at Tsh t_struct_hmac256drbg_context_st initial_state ctx);
             (hmac256drbg_relate initial_state_abs initial_state);
             (data_at Tsh t_struct_mbedtls_md_info info_contents (hmac256drbgstate_md_info_pointer initial_state));
             (Stream s);
             (K_vector kv)
               )
        POST [ tint ]
           EX ret_value:_,
           PROP (
               return_value_relate_result (mbedtls_HMAC256_DRBG_generate_function s initial_state_abs out_len contents) ret_value
             )
           LOCAL (temp ret_temp ret_value)
           SEP (
             (match mbedtls_HMAC256_DRBG_generate_function s initial_state_abs out_len contents with
                | ENTROPY.error _ _ => (data_at_ Tsh (tarray tuchar out_len) output)
                | ENTROPY.success (bytes, _) _ => (data_at Tsh (tarray tuchar out_len) (map Vint (map Int.repr bytes)) output)
              end
             );
             (hmac256drbgabs_common_mpreds (hmac256drbgabs_generate initial_state_abs s out_len contents) initial_state ctx info_contents);
             (data_at Tsh (tarray tuchar add_len) (map Vint (map Int.repr contents)) additional);
             (Stream (get_stream_result (mbedtls_HMAC256_DRBG_generate_function s initial_state_abs out_len contents)));
             (K_vector kv)
           ).
\end{lstlisting}

The main difference between this specification and the API specification of \lstinline{reseed} is that the generate function takes as input a pointer (\lstinline{output}) to a buffer of size \lstinline{out_len}, which is filled with the resulting stream of bytes on success and left alone on failure.
